---
title: "Final Project: Progress Report"
date: today

output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

# Overview

> In this progress report, you'll show some intermediate results of your final project. (Note: This milestone is considered as part of the project management. The grades are only tentative. You should focus on getting some progress. Your final project outcome will outweight the intermediate results.)

0. (5%) Fill the basic information

    * Project title: {replace this with your project title}
    * Repository: {replace this with your git repository link}
    * Team member(s): {replace the following with your member information}

1. (40%) Extend your abstract, describe your project in more details. It should be 300--500 words in length providing:
    + your project goal, or the problem you plan to work on; 
    + (motivation and significance) why the problem is interesting and/or important; 
    + the approach you plan to take, including what data mining tasks you will perform, and what potential techniques you will try; 
    + what dataset you plan to use and how you will get the data (if the data is publicly available, provide the exact reference to the data; otherwise, provide a description about the data source).

2. (30%) Give some preliminary description or analysis about your dataset(s). You should have early numbers or a figure in this report. This part can be short or long, depending on your actual progress. 

3. (25%) The following questions are design to help you manage the progress in doing the final project. Your answers don't need to be long or detailed but will show that you have a plan to address them in your final report.
    a) What do you try to accomplish in this project? What have you done so far?
    b) What are the strengths/novelty of your proposed idea? Why is the problem challenging?
    c) How will you evaluate your method(s)? What are the performance measures and baseline methods?
    d) Have you found any the research or works related to your project/problem/data? Where do you find the related work? 
    e) Are there any challenges you encounter so far? How do you plan to solve it?


```{r document_setup, echo=F, message=F, warning=F}
# This chunk can include things you need for the rest of the document
library('ggplot2') ## most of the time you will need ggplot
theme_set(theme_bw()) # change the default ggplot theme to black-and-white

knitr::opts_chunk$set(
  echo=T, ## show your R code chunk
  message = F, ## hide the message
  warning = F, ## hide the warning
  autodep = T ## make sure your separate code chunks can find the dependencies (from other code chunk)
)
```

# 0. Fill the basic information

 - Project title: Biased Against Bias 
 - Repository: https://github.com/tdg24/2935_final_project
 - Team member(s): 
    Gott, Terner (email: tdg24@pitt.edu)
    Xu, Junchao (email: jux16@pitt.edu)


# 1. Extended abstract 

# 2. Preliminary results

We acquired the 2001 Dutch Census dataset from the github repository listed in the paper "A survey on datasets for fairness-aware machine learning". We have done little preliminary analysis at this point. Mostly we have been in the planning phase. This data set is relatively balanced in terms of size. The protected class is sex. So clearly the bias will seek to understand the occupational outcome of men and women with similar backgrounds and home life. From the paper, it appears that economic status, education level, sex and age will have a large impact on the outcome of the model.

```{r}
df <- read.csv("fairness_dataset/Dutch_census/dutch.csv")
print(names(df))

print(xtabs(~sex, data=df))
```



# 3. Your answers to Problem 3.
  a) First in this project we will seek to understand pre, in and post processing techniques that allow researchers to mitigate bias in their data sets and models. We understand that different techniques are best suited to different data sets. The goal of the project is to uncover and address the biases in the data in order to improve upon the results of models in previous works. Clearly this is past data so the biases are in the world. However, we would aim to perform techniques that would have allowed a fairer and more equal prediction of the individuals. 
  
  b) The approach singularly may not contain any novel techniques but the effort entirely may. We are seeking to use many different techniques to mitigate bias in two well studied data sets. The main challenges are two fold. The data is not very recent and therefor subject to the results of a more biased and fractioned society than it is today. Secondly, the data set is a relatively small sample size in comparison with the entirety of people in each of the groups (be it Netherlands population and Law School Students). The challenge becomes sampling or augmenting the data to be an accurate view of the groups at that given time. 